How Drug Usage During Labor and Delivery Affects Infant Health: A Predictive Analysis of Apgar5 Scores

### 2024-11-26 paper/paper.qmd
update paper first draft

The 2023 Natality Dataset provides detailed records of maternal and neonatal health outcomes across diverse populations in the United States. This obervational dataset aggregates data from hospital records, birth certificates, and surveys completed by healthcare providers. Its comprehensive scope includes variables like maternal age, gestational age, socio-economic status, and the application of specific labor-related treatments.

--- The dataset represents a census of live births in the U.S., negating the need for probabilistic sampling. However, for this study, we employed a stratified sampling approach to create a balanced subset. Treatments like steroids and antibiotics often show skewed distributions, with underrepresentation in certain socio-economic or geographic groups. Stratified sampling ensures adequate representation of minority groups, enabling a more equitable evaluation of treatment effects. Simulations using bootstrapped resampling were conducted to validate that the subset retained key characteristics of the full dataset.---

@@

!!![NEED TO WORK MORE] !!!

>>>>

The 2023 Natality Dataset [@rawdata] provides detailed records of maternal and neonatal health outcomes across diverse populations in the United States. This obervational dataset aggregates data from hospital records, birth certificates, and surveys completed by healthcare providers [@natality2023]. Its comprehensive scope includes variables like maternal age, gestational age, socio-economic status, and the application of specific labor-related treatments [@cdc_facility_worksheet_2016].

The natality data involves addressing potential biases inherent in observational datasets. For example, selection bias may arise if certain populations, such as undocumented individuals or those from rural areas, are underrepresented in the data [@tellingstories]. Additionally, measurement errors, such as inconsistencies in self-reported variables like smoking status or prenatal care visits, can affect data reliability [@healthknowledge2024]. To mitigate these challenges, researchers employ weighting techniques to align samples with population distributions and utilize statistical models like propensity score matching to adjust for confounding factors [@Stuart2021].

Integrating survey methodologies with the 2023 US Natality Data enables researchers to enrich findings by combining administrative records with additional information collected through structured questionnaires. This hybrid approach allows for the exploration of causal relationships, such as the effects of prenatal care access on birth outcomes. By adhering to rigorous survey design principles, such as pilot testing and the Total Survey Error framework, and employing robust sampling strategies, researchers can ensure their findings are both internally valid and generalizable. These methods ultimately enhance the utility of natality data for informing public health policies and interventions aimed at improving maternal and child health outcomes.

@@

Observational data serve as the main database in biomedical research, particularly when randomized controlled trials (RCTs) are impractical due to ethical or logistical constraints [@healthknowledge2024]. These datasets, often derived from electronic health records, patient surveys, or longitudinal studies [@healthknowledge2024], provide valuable insights into real-world healthcare outcomes. However, their inherent limitations necessitate careful consideration to ensure robust and credible findings. Among the key challenges are confounding variables, selection bias, and measurement errors [@tellingstories]. For instance, in analyzing the effects of prenatal steroids on neonatal outcomes, unmeasured factors such as maternal health or socioeconomic status can introduce confounding [@natality202]. Advanced statistical techniques like propensity score matching or instrumental variables are essential to isolate causal effects from observational data [@Olier2023ObservationalData].

Another critical issue is selection bias, where certain subgroups (e.g., uninsured individuals or marginalized populations) may be systematically underrepresented in the dataset, thus limiting the generalizability of findings [@tellingstories]. Addressing this requires techniques such as weighting adjustments or sensitivity analyses and, where possible, integration with survey data to fill informational gaps. Additionally, measurement errors and missing data—common in healthcare records—can distort results [@healthknowledge2024]. Strategies such as multiple imputation and rigorous data cleaning are indispensable for mitigating these challenges [@healthknowledge2024]. Coupled with robust ethical safeguards, including informed consent and adherence to privacy standards, these measures ensure that observational studies maintain both scientific and ethical rigor [@healthknowledge2024].

Causal inference remains a central challenge in using observational data, as the same individual cannot simultaneously experience treatment and non-treatment conditions [@tellingstories]. To approximate counterfactuals, researchers often rely on the Neyman-Rubin potential outcomes framework or marginal structural models [@Groves2010]. Complementing this with simulation studies can test the robustness of findings under varying assumptions. By adhering to established frameworks like the Total Survey Error Model and integrating observational datasets with structured surveys, researchers can produce more reliable and actionable insights [@CausalConversations2024]. This approach not only strengthens the internal and external validity [@tellingstories] of the results but also highlights the potential of observational data to inform equitable and effective healthcare interventions.
